---
title: "p-values and normality"
teaching: 30
exercises: 10
questions:
- "FIXME"
objectives:
- "FIXME"
keypoints:
- "FIXME"
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("01-")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# what are we doing when we do a statistical test?

A simple example of doing a statistical test, would be to take two groups, in this case 10 persons, and give them two different sleep inducing drugs. We now measure the increase in the hours they sleep, and try to determine if there is a difference between the two drugs. Ie do they sleep longer taking drug 1 than they do taking drug 2.

This is a classic dataset from the study af two versions (optical isomers) of scopolamine. 

The datset is called "sleep", and is included with R:

```{r}
head(sleep)
```
ID is the patient ID, extra the additional hours of sleep, when taking group (or drug version) 1 or 2.

We can calculate the average number of extra sleep for the two drugs (group):

```{r}
library(tidyverse)
sleep %>% 
  group_by(group) %>% 
  summarise(mean_extra = mean(extra))
```

It would appear that drug 2 provides the test subjects with significant more sleep that drug 1. 

But this might be due to chance! Idling engines or noisy birds might randomly disturb our sleep, and there is a natural variation in how many hours each individual sleeps every night.

Statistical tests will give us a measure of the odds of seeing the difference shown above, purely by change.

Step 1 is to define a NULL-hypothesis, claiming that the diffence is coincidental, and that there is no true difference in the average number of extra hours slept when taking one of the two drugs. 

Step 2, which follows automatically, is that there is a difference. That the difference we are seeing is not due to chance.

One test used for this is the Student's t-test.

Running that test is step 3.

The t-test can be run in two ways. 

Method 1 requires us to have a dataframe with two columns, like we have in the sleep dataset shown above. We have a column with the value we are testing, extra sleep in this case, and a column indicating which group each observation is in.

We can then input the data to the t.test in this manner:

```{r}
test_res_1 <- t.test(extra ~ group, data = sleep)
```
We store the result as the variable test_res. The formual notation extra ~ group can be read as:

The variable extra, as a function of group. Group is a categorical variable, the test subjects take either drug number 1 or number 2.

We need to tell the test which dataframe we are using, specifying "data = sleep".

The alternative way is to extract two vectors, containing the extra hours of sleep, one for each drug:

```{r}
d1 <- sleep[sleep$group==1,]$extra
d2 <- sleep[sleep$group==2,]$extra
```

And then running the t-test in this manner:
```{r}
test_res_2 <- t.test(d1,d2)
```

The two results are not completely identical:

```{r}
test_res_1
```

However the difference is only due to the different ways we inputted the data to the function:
```{r}
test_res_2
```
And what does the output tell us?

We are getting the t-statistics, and the number of degress of freedom. Today they are mostly used for reporting the results, but historically we would take them and use them to look up the p-value in a printed table.

We no longer do that, we get the p-value directly. It is 0.07939.

We are told that the alternative hypothosis is that the true difference in means for our two samples is not equal to 0. Which is a cumbersome way of saying that the alternative hypothesis is that the means are different.

And what is the conclusion? 

The p-value tells us the probability that if the two sam


Resultat?
t er den testværdi der beregnes ud fra data. Den dukker ofte op, også i andre tests. Og undertiden hedder den noget andet end t. df er antallet af frihedsgrader. (DET MÅ VI HELLERE FORKLARE ET ELLER ANDET STED OGSÅ).
Dem bruger vi sjælden tid på, men det er ofte en god ide at rapportere dem når man skriver opgaver og den slags.

Den vi typisk fokuserer på er p-værdien. Den finder vi ud fra antallet af frihedsgrader og t-værdien, og i dette tilfælde er den 0.07939.

Hvad betyder det? Det betyder, at vi ser en forskel på de to middelværdier. Den er på 1.58. Og vi har en NULL-hypotese om at der ikke er forskel.

Hvis der ikke er forskel på de to middelværder. Så er sandsynligheden for at vi ser en forskel på dem, der er så ekstrem at den er på 1.58, 0.07839414.

Eller, der er næsten 8% chance for, at hvis der ingen forskel er, at vi ser en forskel på 1.58 timer mellem de to grupper.

Det er en relativt stor chance. Typisk er vores succeskriterium 5%. 
Så vi kan IKKE afvise NULL-hypotesen. Vi kan ikke, med sikkerhed, sige at der er en forskel.



# Frihedsgrader - hvad er det?

# normalitet

Cushny, A. R. and Peebles, A. R. (1905) The action of optical isomers: II hyoscines. The Journal of Physiology 32, 501–510. (issue 5-6)
doi: 10.1113/jphysiol.1905.sp001097

Student (1908) The probable error of the mean. Biometrika, 6, 20.
side 1 til 25, hæfte 1 vol 6

{% include links.md %}