---
title: "Summary statistics"
teaching: 30
exercises: 10
questions:
- "FIXME"
objectives:
- "FIXME"
keypoints:
- "FIXME"
source: Rmd
---


```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("02-")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Summary statistics 

Summary statistics is a collective term for descriptive summarisations of observations. 

They are typically divided into measurements of:
Location
dispersion
Shape
Dependence



## Location

Looking at our data, where are they?


### Mean

The average or geometric mean, but usually just mean, in this case the 
average weight of penguins:
```{r}
mean(penguins$body_mass_g, na.rm = T)
```
Note that we are asking the mean function to remove missing, NA, values before
calculating the mean.

### Median

Placing all our penguins in order from the lightest to the heaviest, what is the
weight of the middle penguion?

```{r}
median(penguins$body_mass_g, na.rm = T)
```
That the median penguin is lighter than the average penguin tells us that weight 
amongst these penguins is skewed towards lighter, smaller penguins.


### Mode

Mode is the most frequent or common observation of a given variable. 

R does not have a built-in standard function for this. There are several ways of 
getting it, this is just one way of getting the mode of the body mass of the 
penguins:
```{r}
penguins %>%  
  group_by(body_mass_g) %>% 
  summarise(number = n()) %>% 
  arrange(desc(number)) %>% 
  slice(1) %>% 
  pull(body_mass_g)
```
We group the penguins by body_mass_g, and get the number of observations in 
each group of penguins. Then we arrange the resulting dataframe in order of 
descending number of observations, use the slice function to give us only the
first row of data in that dataframe. And finally we use the pull function to pull
out the value in the column body_mass_g.

### Interquartile mean

Lidt mere bøvlet...

## Spread

Given the location of our data, how much are they spread out around that location

SD, 
varians, 
range, 
inter quartile range, 
absolute deviation, 
mean absolute difference, 
distance standard deviation


## Shape

When we look at hte distribution of our data, what shape do they have. Summarised to numbers.

Skewness

kurtoses

## Dependency

If we measure more than one variable, are there dependencies between them. 
Pearson product-moment correlation coefficient

# quick summarystatistics


```{r}
summary(penguins$bill_length_mm)
```
```{r}
boxplot(penguins$bill_depth_mm)
```
More often plotted as a series of boxplots where we make one boxplot for each of a
number of groups.

# details:

## Location

Getting the mean:
Looking at our data, where are they?

mean,
median
mode
interquartile mean

## Spread

Given the location of our data, how much are they spread out around that location

SD, varians, range, inter quartile range, absolute deviation, mean absolute difference, distance standard deviation


## Shape

When we look at hte distribution of our data, what shape do they have. Summarised to numbers.

Skewness, kurtoses

## Dependency

If we measure more than one variable, are there dependencies between them. 
Pearson product-moment correlation coefficient

{% include links.md %}