---
title: "Identify textual changes"
output: html_document
---

# Identify textual changes

As example data we use the first chapter of two different editions of the novel *Frankenstein*.

Metadata first book:

- **Author:** Shelley, Mary Wollstonecraft, 1797–1851
- **Title:** *Frankenstein; Or, The Modern Prometheus*
- **Original publication:** United Kingdom: Lackington, Hughes, Harding, Mavor, & Jones, 1818.

Metadata second book:

- **Author:** Shelley, Mary Wollstonecraft, 1797–1851
- **Title:** *Frankenstein; Or, The Modern Prometheus*
- **Original publication:** United Kingdom: H. Colburn and R. Bentley, 1831.

After we have extracted the first chapter of each book, we try two different approaches to text comparison.

For the first approach we use the R package `diffobj`. Its `diffChr()` function compares character vectors and provides a visual representation of the differences between them.

For the second approach we use Levenshtein distance to calculate the similarity between the two texts. R provides Levenshtein distance directly through `adist()`.

# Download the two novels from Gutenberg.org

Download the two novels from Gutenberg.org and extract the first chapter of each book.

HER SKAL VI NOK HAVE NOGET MED GUTENBERGR PAKKEN I STEDET.

```{r}
library(gutenbergr)
# Get the 1818 edition
url1 <- "https://gutenberg.org/cache/epub/41445/pg41445.txt"

raw_text1 <- readLines(
  url1,
  warn = FALSE,
  encoding = "UTF-8"
) |>
  paste(collapse = "\n")

# Get Chapter I
chap1_1 <- sub(
  "^[\\s\\S]*?(CHAPTER I[\\s\\S]*?)CHAPTER II[\\s\\S]*$",
  "\\1",
  raw_text1,
  perl = TRUE
)


# Get the 1831 edition
url2 <- "https://gutenberg.org/cache/epub/42324/pg42324.txt"

raw_text2 <- readLines(
  url2,
  warn = FALSE,
  encoding = "UTF-8"
) |>
  paste(collapse = "\n")

# Get Chapter I
chap1_2 <- sub(
  "^[\\s\\S]*?(CHAPTER I[\\s\\S]*?)CHAPTER II[\\s\\S]*$",
  "\\1",
  raw_text2,
  perl = TRUE
)
```

The regular expression keeps the text beginning with `CHAPTER I` and ending immediately before `CHAPTER II`.

# Preprocess the text to get homogeneous data

We define a small preprocessing function. It:

1. replaces carriage returns and line breaks with spaces,
2. removes characters other than letters, numbers, underscores, whitespace and full stops,
3. reduces repeated whitespace to a single space.

```{r}
preprocess_text <- function(text) {

  text <- gsub(
    "[\r\n]",
    " ",
    text
  )

  text <- gsub(
    "[^[:alnum:]_[:space:].]",
    "",
    text
  )

  text <- gsub(
    "\\s+",
    " ",
    text,
    perl = TRUE
  )

  trimws(text)
}


edition1818 <- preprocess_text(chap1_1)
edition1831 <- preprocess_text(chap1_2)
```

# Using `diffobj` for text comparison

The Python version uses `difflib.Differ()`. In R, the package `diffobj` provides similar functionality.

Install the package if necessary:

```{r eval=FALSE}
install.packages("diffobj")
```

Load it:

```{r}
library(diffobj)
```

We first make a function that splits the texts into sentences and compares them.

```{r}
compare_texts <- function(text1, text2) {

  # Split the texts into sentences
  lines1 <- strsplit(
    text1,
    "\\.\\s+",
    perl = TRUE
  )[[1]]

  lines2 <- strsplit(
    text2,
    "\\.\\s+",
    perl = TRUE
  )[[1]]

  # Compare the texts
  diffChr(
    lines1,
    lines2,
    pager = "off"
  )
}
```

Compare the first 2000 characters in the first chapter of each edition:

```{r}
compare_texts(
  substr(edition1818, 1, 2000),
  substr(edition1831, 1, 2000)
)
```

`diffChr()` highlights text that occurs only in the first text, text that occurs only in the second text, and text shared by the two versions.

The precise visual representation depends on the output device. In an interactive R session, `diffobj` can use colour and formatting to make additions and deletions easier to see.

# Using Levenshtein distance to compute the similarity of two texts

The Python example uses `fuzzywuzzy` and its `fuzz.ratio()` function.

R already contains a function for calculating Levenshtein edit distance: `adist()`.

The Levenshtein distance is the minimum number of insertions, deletions and substitutions required to transform one string into another.

We can turn the distance into a percentage similarity score.

```{r}
compare_sentences <- function(sentence1, sentence2) {

  # Calculate Levenshtein distance
  distance <- adist(
    sentence1,
    sentence2
  )[1]

  # Convert the distance to a similarity ratio
  max_length <- max(
    nchar(sentence1),
    nchar(sentence2)
  )

  ratio <- 100 * (
    1 - distance / max_length
  )

  ratio
}
```

Calculate the similarity between the complete first chapters:

```{r}
similarity <- compare_sentences(
  edition1818,
  edition1831
)

sprintf(
  "Similarity: %.0f%%",
  similarity
)
```

The result is a normalized Levenshtein similarity score:

\[
\text{similarity}
=
100
\left(
1 -
\frac{\text{Levenshtein distance}}
     {\text{length of the longest text}}
\right)
\]

A score of 100% means that the two strings are identical. Lower scores indicate increasingly large differences.

# Complete example

The complete workflow can therefore be written compactly as:

```{r eval=FALSE}
library(diffobj)

url1 <- "https://gutenberg.org/cache/epub/41445/pg41445.txt"
url2 <- "https://gutenberg.org/cache/epub/42324/pg42324.txt"

raw_text1 <- readLines(
  url1,
  warn = FALSE,
  encoding = "UTF-8"
) |>
  paste(collapse = "\n")

raw_text2 <- readLines(
  url2,
  warn = FALSE,
  encoding = "UTF-8"
) |>
  paste(collapse = "\n")


chap1_1 <- sub(
  "^[\\s\\S]*?(CHAPTER I[\\s\\S]*?)CHAPTER II[\\s\\S]*$",
  "\\1",
  raw_text1,
  perl = TRUE
)

chap1_2 <- sub(
  "^[\\s\\S]*?(CHAPTER I[\\s\\S]*?)CHAPTER II[\\s\\S]*$",
  "\\1",
  raw_text2,
  perl = TRUE
)


preprocess_text <- function(text) {

  text <- gsub(
    "[\r\n]",
    " ",
    text
  )

  text <- gsub(
    "[^[:alnum:]_[:space:].]",
    "",
    text
  )

  text <- gsub(
    "\\s+",
    " ",
    text,
    perl = TRUE
  )

  trimws(text)
}


edition1818 <- preprocess_text(chap1_1)
edition1831 <- preprocess_text(chap1_2)


compare_texts <- function(text1, text2) {

  lines1 <- strsplit(
    text1,
    "\\.\\s+",
    perl = TRUE
  )[[1]]

  lines2 <- strsplit(
    text2,
    "\\.\\s+",
    perl = TRUE
  )[[1]]

  diffChr(
    lines1,
    lines2,
    pager = "off"
  )
}


compare_texts(
  substr(edition1818, 1, 2000),
  substr(edition1831, 1, 2000)
)


compare_sentences <- function(sentence1, sentence2) {

  distance <- adist(
    sentence1,
    sentence2
  )[1]

  max_length <- max(
    nchar(sentence1),
    nchar(sentence2)
  )

  100 * (
    1 - distance / max_length
  )
}


similarity <- compare_sentences(
  edition1818,
  edition1831
)

sprintf(
  "Similarity: %.0f%%",
  similarity
)
```
