---
title: "Prompting in text analysis"
output: html_document
---

# Prompting in text analysis

This document gives examples of how prompt engineering can be used as a method in text analysis.

You define a task, the model reads the text and returns an answer that you can work with further.

We work with the fairy tale *The Tinderbox* by H. C. Andersen.

# Using Mistral from R

In R we can use the [`ellmer`](https://ellmer.tidyverse.org/) package to communicate with large language models.

`ellmer` supports a number of providers, including Mistral, and provides the function `chat_mistral()`.

Install the package once:

```{r eval=FALSE}
install.packages("ellmer")
```

Load it:

```{r}
library(ellmer)
```

## API key

To use the Mistral API you need a Mistral API key.

Do not save the key directly in an R Markdown document or an R script that you intend to share.

Instead, save it as the environment variable:

```text
MISTRAL_API_KEY
```

One convenient way to do this is to edit your `.Renviron` file:

```{r eval=FALSE}
usethis::edit_r_environ()
```

Add a line such as:

```text
MISTRAL_API_KEY=your_key_here
```

Save the file and restart R.

`ellmer::chat_mistral()` will automatically use the key stored in `MISTRAL_API_KEY`.

# Create a Mistral chat

A minimal Mistral chat can be created like this:

```{r eval=FALSE}
chat <- chat_mistral(
  model = "mistral-small-latest"
)

chat$chat(
  "What is the capital of Denmark?"
)
```

The `chat` object represents a conversation.

This means that it remembers previous messages:

```{r eval=FALSE}
chat <- chat_mistral(
  model = "mistral-small-latest"
)

chat$chat(
  "My favourite animal is the wombat."
)

chat$chat(
  "What is my favourite animal?"
)
```

For the exercises below, however, we want each analysis to be independent of the previous analyses.

We therefore define a small helper function that creates a new chat for every request.

```{r}
ai_assistent <- function(
  prompt,
  system_prompt = "You are a helpful assistant.",
  temperature = 0,
  max_tokens = NULL
) {

  chat <- chat_mistral(
    model = "mistral-small-latest",
    system_prompt = system_prompt,
    params = params(
      temperature = temperature,
      max_tokens = max_tokens
    ),
    echo = "none"
  )

  chat$chat(prompt)
}
```

The function has four arguments:

- `prompt`: the user prompt
- `system_prompt`: general instructions to the model
- `temperature`: controls variation in the generated response
- `max_tokens`: optionally limits the length of the generated response

# Load text: *The Tinderbox*

We load the fairy tale from a text file.

```{r}
the_tinderbox <- readLines(
  "the tinderbox.txt",
  warn = FALSE,
  encoding = "UTF-8"
) |>
  paste(collapse = "\n")
```

Look at the first 100 characters:

```{r}
substr(
  the_tinderbox,
  1,
  100
)
```

# Ask for a summary

We can insert the fairy tale directly into the prompt.

```{r}
prompt <- paste0(
  "What is the text about?\n\n",
  the_tinderbox
)

summary <- ai_assistent(prompt)
```

Show the answer:

```{r}
cat(summary)
```

Save the answer to a text file:

```{r}
writeLines(
  summary,
  "summary_tinderbox.txt",
  useBytes = TRUE
)
```

Look at your summary.

Are you satisfied with the answer?

Should the task be formulated more precisely?

# Prompt settings

Several settings can influence the model's response.

## Model

We explicitly selected:

```{r eval=FALSE}
model = "mistral-small-latest"
```

Explicitly selecting a model makes the analysis less dependent on changes to the provider's default model.

## User prompt

The user prompt defines the concrete task.

For example:

```text
You are presented with a fairy tale.
Write a summary.

[insert fairy tale]
```

or:

```text
You are presented with a fairy tale.
Return the main characters as bullet points.

[insert fairy tale]
```

In our function this is the `prompt` argument.

## System prompt

The system prompt gives the model general instructions about how it should behave.

For example:

```text
You work with literary analysis.
Always answer in English.
Never return more than five points.
```

With `ellmer`, the system prompt is supplied when the chat is created:

```{r eval=FALSE}
chat <- chat_mistral(
  model = "mistral-small-latest",
  system_prompt = "You work with literary analysis."
)
```

## Temperature

Temperature controls variation in the generated response.

For analytical tasks we often want relatively stable responses, so we use a low temperature:

```{r eval=FALSE}
params(
  temperature = 0
)
```

Higher values generally allow more variation in the generated text.

## Maximum tokens

We can also limit the maximum number of tokens generated:

```{r eval=FALSE}
params(
  temperature = 0,
  max_tokens = 1500
)
```

A token is not necessarily the same as a word. A word may consist of one or more tokens.

# A focused literary analysis

We now make the task more specific.

First define a system prompt:

```{r}
system_content <- paste(
  "You work with literary analysis.",
  "Answer in English.",
  "Never return more than five points.",
  "",
  "Divide each point into:",
  "1. an example from the text.",
  "2. a brief paraphrase of the example.",
  "3. an interpretation that justifies the example.",
  sep = "\n"
)
```

Then create the user prompt:

```{r}
prompt <- paste0(
  "You are presented with a fairy tale by H. C. Andersen between <text> tags.\n\n",
  'Your task is to find examples of "animals as helpers".\n\n',
  "<text>\n",
  the_tinderbox,
  "\n</text>"
)
```

Run the analysis:

```{r}
response <- ai_assistent(
  prompt,
  system_prompt = system_content,
  temperature = 0,
  max_tokens = 1500
)
```

Show the answer:

```{r}
cat(response)
```

Save it:

```{r}
writeLines(
  response,
  "analysis_animals_tinderbox.txt",
  useBytes = TRUE
)
```

# A structured analysis

A prompt can contain several tasks and specify how the answer should be formatted.

First define the system prompt:

```{r}
system_content <- paste(
  "You work with literary analysis.",
  "Always answer in English.",
  "Stay close to the text of the fairy tale.",
  "Do not invent information that is not supported by the text.",
  "Do not use an introduction.",
  sep = "\n"
)
```

Now create the prompt:

```{r}
prompt <- paste0(
  "
Your task is to perform the following actions:

1 - Summarise the following text, delimited by <text> tags, in four sentences.
2 - Return a list containing the fairy tale's \"good\" characters.
3 - Return a list containing the fairy tale's \"evil\" characters.
4 - Return a list containing the magical objects.
5 - Return a short account of the fairy tale's repetition structure.

Use the following format:

Summary: see 1 above.
Good characters: see 2 above.
Evil characters: see 3 above.
Magical objects: see 4 above.
Repetition structure: see 5 above.

<text>
",
  the_tinderbox,
  "
</text>"
)
```

Send it to Mistral:

```{r}
response <- ai_assistent(
  prompt,
  system_prompt = system_content,
  temperature = 0,
  max_tokens = 1500
)
```

Show the answer:

```{r}
cat(response)
```

Save the result:

```{r}
writeLines(
  response,
  "analysis2_tinderbox.txt",
  useBytes = TRUE
)
```

# Find and interpret semantic fields

A language model can also help identify passages relevant to a particular question.

Below we first use the model to find two passages and then classify their sentiment.

# Find two passages

Define the system prompt:

```{r}
system_content <- paste(
  "You work with literary analysis.",
  "Answer only on the basis of the supplied text.",
  "Do not use introductions or other explanations in your response.",
  sep = "\n"
)
```

Create the prompt:

```{r}
prompt <- paste0(
  "
Find the two passages of the fairy tale where the soldier meets the witch.

1st time: where he meets the witch.
2nd time: where he kills the witch.

Return the two passages from the English version as text.

Use the following format:

1. first passage.
#####
2. second passage.

Fairy tale:

<text>
",
  the_tinderbox,
  "
</text>"
)
```

Send the prompt:

```{r}
response <- ai_assistent(
  prompt,
  system_prompt = system_content,
  temperature = 0,
  max_tokens = 1500
)
```

Show the answer:

```{r}
cat(response)
```

Save the passages:

```{r}
writeLines(
  response,
  "soldier_and_witch.txt",
  useBytes = TRUE
)
```

# Sentiment analysis of the passages

Load the two extracted passages:

```{r}
soldier_and_witch <- readLines(
  "soldier_and_witch.txt",
  warn = FALSE,
  encoding = "UTF-8"
) |>
  paste(collapse = "\n")
```

Create a prompt asking the model to classify the sentiment of each passage:

```{r}
prompt <- paste0(
  "
You are presented with two different passages.

What is the sentiment of each passage?

Choose between:

- positive
- negative
- neutral

Passages:

<passages>
",
  soldier_and_witch,
  "
</passages>

Use the following format:

1. first passage: [sentiment]. Justification.
2. second passage: [sentiment]. Justification.
"
)
```

Run the analysis:

```{r}
response <- ai_assistent(
  prompt,
  system_prompt = "You work with text analysis.",
  temperature = 0,
  max_tokens = 1500
)
```

Show the result:

```{r}
cat(response)
```

Save it:

```{r}
writeLines(
  response,
  "soldier_and_witch_sentiment.txt",
  useBytes = TRUE
)
```

# Working directly with `ellmer`

Our `ai_assistent()` function is convenient because every call is independent.

Sometimes, however, we deliberately want an ongoing conversation.

Then we can work directly with the `Chat` object:

```{r eval=FALSE}
chat <- chat_mistral(
  model = "mistral-small-latest",
  system_prompt = "You work with literary analysis.",
  params = params(
    temperature = 0
  )
)

chat$chat(
  paste0(
    "Summarise this fairy tale:\n\n",
    the_tinderbox
  )
)

chat$chat(
  "Now list the three most important characters."
)

chat$chat(
  "Which of those characters change during the story?"
)
```

Each new call is interpreted in the context of the previous conversation.

That is useful for an interactive dialogue, but less useful when we want separate, reproducible text-analysis tasks.

# A reusable helper script

The helper function can be saved in a separate file called, for example, `mistral.R`:

```{r eval=FALSE}
library(ellmer)

ai_assistent <- function(
  prompt,
  system_prompt = "You are a helpful assistant.",
  temperature = 0,
  max_tokens = NULL
) {

  chat <- chat_mistral(
    model = "mistral-small-latest",
    system_prompt = system_prompt,
    params = params(
      temperature = temperature,
      max_tokens = max_tokens
    ),
    echo = "none"
  )

  chat$chat(prompt)
}
```

Then load it in another analysis with:

```{r eval=FALSE}
source("mistral.R")
```

and use:

```{r eval=FALSE}
answer <- ai_assistent(
  "Your prompt here"
)
```

# Complete minimal example

The essential workflow is now very small:

```{r eval=FALSE}
library(ellmer)

text <- readLines(
  "the tinderbox.txt",
  warn = FALSE,
  encoding = "UTF-8"
) |>
  paste(collapse = "\n")

chat <- chat_mistral(
  model = "mistral-small-latest",
  system_prompt = "You work with literary analysis.",
  params = params(
    temperature = 0
  )
)

answer <- chat$chat(
  paste0(
    "Summarise the following fairy tale in five sentences:\n\n",
    text
  )
)

cat(answer)
```
